Local PCA Shows How Population Structure Differs Along the Genome

### 2016-05-25 writeup/local_pca_paper.tex
update
peter rewrote introduction

Dimension reduction techniques,
such as principal components,
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship this uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find substantial variation on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlates with local gene density,
and may be caused by background selection or local adaptation.

@@

The kinship matrix contains the kinship coefficient for pairwise individuals. 
Kinship coefficient defines the genetic relatedness between individuals.
It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
Populations are often structured in some way while there are systematic genetic variation between populations.

About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
PCA results are derived from the covariance matrix of genotype matrix. 
The results of PCA can be directly related to the underlying genealogical history of the samples, 
such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

PCA is often used in genome-wide association studies (GWAS) for stratification correction \citep{price2006principal}. 
However, different parts of genome have different genetic features. 
First, each site of DNA may have different gene tree. 
The covariance matrix of genotype matrix averages those gene trees. 
For a genomic region, if individuals have alleles closer in gene tree, they tends to be more close in PC projections for that region. 
Different DNA segments may have different gene tree and therefore different population structure for those segments. 
Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
Selective sweeps cause local recent ancestry or short trees.
Balancing selection causes deep trees.
Background selection causes shallow ones.
Third, if a chromosome inversion is polymorphic in the sample, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
Other effects, like noise, introgression might also influence population structure.

Investigating the genomic variation along the genome can help us to have a
better understanding of the relation between genome structure and population
structure, and could possibly lead to more powerful methods for GWAS.

To investigate this, we cut each chromosome into windows (with hundreds to
thousands of SNPs in each), applied PCA to each window, and visualized how
population structure, as summarized by PCA, varies along windows.

In this project, we used SNP data for human, \textit{Medicago truncatula}, and
whole genome sequencing data for \textit{Drosophila}.  Based on the principal
components, we can estimate the similarity of population structure contained in
each genome window.  To quantify similarity of population structure between
windows, we constructed for each window an approximate, scaled covariance
matrix based on the first two PCs measured the pairwise Euclidean distance
between those matrices.  We use multidimensional scaling to visualize the
relationships between windows, which reduces the pairwise distance matrix to
lower dimension while preserving the distance information between windows as
well as possible \citep{borg2005modern}.  To interpret the results of MDS, we
combine known genome feature information for each species, such as the
distribution of inversions, and heterochromatin and gene density along the
genome. Each species showed distinct patterns, reflecting differences in their
biology.

\plr{Improve this paragraph:}
Other methods for visualizing population structure are like STRUCTURE,
\citep{pritchard2000inference,falush2003inference,
falush2007inference,hubisz2009inferring} 
model-based approach,
\citep{yang2012modelbased} 
maps of heterozygosity
\citep{ramachandran2005support}.

A number of methods for dimensionality reduction also use a strategy of ``local PCA''
\citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
performing PCA not on the entire dataset but instead on subsets of observations,
providing ``local'' pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although the ultimate goals and methods for visualization are different.
Future methods for visualization of genomic data
may benefit from other advances in this substantial literature \citep[reviewed in][]{vandermaaten2009dimensionality}.

\section{Other Introduction}

Catchy start: the kinship matrix goes back to almost Mendel and is essential in GWAS;
however, it is well-known that actual relatednesses have a lot of noise about the expected value,
and depend on where on the genome you look;
this is why scans for selective sweeps work.

Review of kinship matrix: 
it's either an expected kinship, given the pedigree;
or an estimated genome-wide average.
Wright's path coefficients \citep{wright1943isolation}.
Why it helps with confounding for GWAS.
Graham \& Vince's paper, maybe.
Like IBD, is only well-defined in a known pedigree, up to the founders.
Kinship and confounding reviewed in \citet{astle2009population}.

Kinship matrices differ for sex chromosomes and the like.

Review of selection causing differential patterns along the genome.
Locally everything is treelike (gene trees);
kinship matrix is an average of these (write equation for this).
Selective sweeps cause local recent ancestry/short trees.
Balancing selection causes deep trees.
Background selection, shallow ones.
Extreme examples of free gene flow in some places between species: e.g. Heliconius.
Introgression may be nonuniform, e.g. neanderthal, others.
Refs: hitchhiking \citep{maynardsmith1974hitchhiking},
\citet{kim2011hitchhiking} study hitchhiking in a spatially subdivided population,
\citet{mcvean2007structure} looks at the effect of selection on LD.
\citet{barton2000genetic} reviews hitchhiking.
\citet{bierne2010distinctive} discusses how hitchhiking effect decreases with geographic distance.
\citet{charlesworth2003review} reviews patterns of diversity, relating spatial structure to effects of selection.

Review of methods looking along the genome:
argweaver, HMM between species, ???

What is "population structure"?
Asks which "populations" are closely related, more diverged, how much diversity do they harbor.
Often geographic.
Vital in exploratory data analysis.
It is a summary of kinship: 
lack of migration between pops causes a deficit in connections through the pedigree,
and so affects kinship.
Wright defined $F_{ST}$ in \citep{wright1949genetical}, and says ``It has probably occurred to the reader that the coefficientof inbreeding may mean very different
things in different cases.''

Review of methods for visualizing pop structure:
PCA, structure \citep{falush2003inference}, EEMS, \citep{petkova2014visualizing}, \citep{yang2012modelbased},
Maps of heterozygosity \citep{ramachandran2005support}.
Genealogical interpretation of PCA by \citet{mcvean2009genealogical}.
Other semi-related stuff:
estimation of covariance matrices;
local pca(?);

@@

The phrase ``population structure'' refers to reduced gene flow between subpopulations, often because of geographical isolation.
This leads to systematic patterns in genome-wide mean kinship,
and so visualizations of kinship are said to depict population structure,
rather than depicting the \emph{effects} of population structure (which would be more accurate).
However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome \citep{introgression_review},
and patterns of polymorphism and divergence vary significantly depending on factors including local gene density \citep{linked_selection,sattath}.
This implies that the population structure of a species depends on which part of the genome is being examined.

>>>>

Dimension reduction techniques,
such as principal components,
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship this uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find substantial variation on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is a useful addition to the exploratory toolbox
of population genomics.

@@

% The kinship matrix contains the kinship coefficient for pairwise individuals. 
% Kinship coefficient defines the genetic relatedness between individuals.
% It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
% It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
% It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
% Populations are often structured in some way while there are systematic genetic variation between populations.

The phrase ``population structure'' refers to reduced gene flow between subpopulations, 
often because of geographical isolation.
However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
and patterns of polymorphism and divergence can vary significantly depending on factors including local gene density.
This implies that, paradoxically, 
the population structure of a species depends on which part of the genome is being examined.

Population structure leads to systematic patterns in genome-wide mean kinship,
and so it is said that visualizations of kinship depict population structure,
rather than the \emph{effects} of population structure (which would be more accurate).
The \emph{kinship coefficient} for a pair of individuals
gives the expected proportion of their genome
that the two have inherited identically by descent;
for a single individual it is the inbreeding coefficient.
However, as \citet{wright1949genetical} wrote,
``It has probably occurred to the reader that the coefficient of inbreeding
may mean very different things in different cases.''
The kinship coefficient originally referred to the expected probability
of coinheritance within a given pedigree;
while modern applications to ``unrelated'' individuals
use a genetic covariance matrix to estimate the \emph{realized} proportion of the genome
coinherited from sufficiently recent ancestors.

Realized kinship summarizes the shapes
of the genealogical that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
and therefore patterns of kinship.
Indeed,
chromosome-scale variation in diversity and divergence has been observed in many species
\citep[e.g.][]{sattath2011pervasive};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citet{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor what the major axes of variation are.

Patterns in genome-wide kinship are often summarized
by applying principal components analysis (PCA) \citep{patterson2006population} 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
% PCA results are derived from the covariance matrix of genotype matrix. 
The results of PCA can be directly related to the underlying genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
These patterns are often geographical, producing ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes},
although other patterns emerge if recent migration or nongeographic kinship patterns are important
\citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

As reviewed by \citet{astle2009population}, 
modeling ``background'' kinship between samples
is essential to successful genome-wide association studies,
and PCA has often been used for stratification correction \citep{price2006principal},
and so investigating variation in kinship along the genome can help us to have a
better understanding of the relation between genome structure 
and population structure, possibly leading to more powerful methods for association studies.

To investigate how population structure varies along the genome in several datasets, 
we cut each genome into windows (with hundreds to thousands of SNPs in each), 
applied PCA to each window, 
and visualized the major ways that population structure, as summarized by PCA, varies among windows.
% In this project, we used SNP data for human, \textit{Medicago truncatula}, and
whole genome sequencing data for \textit{Drosophila}.  
% Based on the principal components, we can estimate the similarity of population structure contained in each genome window.  
To quantify similarity of population structure between
windows, we constructed for each window an approximate, scaled covariance
matrix based on the first few principal components, and measured the pairwise Euclidean distance
between those matrices.  
We then use multidimensional scaling to visualize the
relationships between windows, which reduces the pairwise distance matrix to
lower dimension while preserving the distance information between windows as
well as possible \citep{borg2005modern}.

% To interpret the results of MDS, we combine known genome feature information for each species, such as the
% distribution of inversions, and heterochromatin and gene density along the genome. 
Each species showed distinct patterns, reflecting differences in their biology;
before presenting results,
it may help to have a prior idea of what factors are expected to affect population structure,
and how.
PCA summarizes patterns in kinship found in the genetic covariance matrix,
which is an average across locus-specific genealogies.
% However, different parts of genome have different genetic features. 
% First, each site of DNA may have different gene tree. 
% The covariance matrix of genotype matrix averages those gene trees. 
If individuals are closer in the genealogies of a given genomic region, 
they tends to be more close in the PCA maps for that region. 
% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
% Selective sweeps cause local recent ancestry or short trees.
% Background selection causes shallow ones.
Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
A genomic region with many targets for selection experiences
background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
which would tend to generally shorten genealogical trees in the region,
similar to a reduction in effective population size \citep{hudson1995deleterious}.
% Balancing selection causes deep trees.
On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
and population structure locally describes which individuals have which alleles
rather than geographical proximity.
Finally, 
since recombination is suppressed between opposite orientations of a chromosomal inversion
near its breakpoints,
genealogies in these regions separate samples carrying the two orientations of the inversion.
In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
as a way to identify inversions).
% if a chromosomal inversion is polymorphic in the sample, 
% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints. Recombination suppression across inversions thus results in different genome structure and population structure. 
% Other effects, like noise, introgression might also influence population structure.
We will look for strong variation in population structure shared across large regions of the genome.
Many of the effects listed above (such as single selective sweeps or inversions) 
are not expected to have similar effects on population structure in different regions of the genome
because of randomness in which samples end up in which group;
but if these coincide with a region of reduced recombination (such as an inversion),
these could drive major patterns of variation.
The more subtle effects of genome-wide linked selection could be shared across large regions
due to large-scale variation in gene density
(as by background selection or local adaptation at many genes).

We chose PCA to summarize population structure,
but other methods, such as STRUCTURE \citep{falush2003inference},
SPA \citep{yang2012modelbased},
SpaceMix \citep{bradburd2015spatial},
or other matrix factorization methods \citep{engelhardt2010analysis}
would highlight different sources of variation in the data.
% There are other methods for visualizing population structure are like STRUCTURE,
% \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} 
% model-based approach,
% maps of heterozygosity
% \citep{ramachandran2005support}.

Finally, a number of general methods for dimensionality reduction also use a strategy of ``local PCA''
\citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
performing PCA not on the entire dataset but instead on subsets of observations,
providing ``local'' pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although the ultimate goals and methods for visualization are different.
Future methods for visualization of genomic data
may benefit from other advances in this substantial literature 
\citep[reviewed in][]{vandermaaten2009dimensionality}.

@@

(sentence about how we found variation in pop structure)
